Basic Model formulation

### 2015-01-27 documentation/model_formulation.Rmd
update model formulation

defplot <- function(...) plot(..., 
            type = "l", 
            bty = "l", cex = 0.7, las = 1, 
            family = "Times",
            xlim = c(0,1), ylim = c(0,1), 
            xaxs = "i", yaxs = "i", 
            xaxp = c(0,1,2), yaxp = c(0,1,2), 
            xlab = "vegetation cover")

@@

$$ G(V) = rV(1-V/K) \,,$$

@@

```{r, echo=FALSE}
G <- function(P, r = 2, K = 0.9) r*P*(1-P/K)

@@

@@

$$ G(V) = r_0 V^{a+1} (1-V/K) \,,$$

@@

G <- function(P, r_0 = 2, a = 2, K = 0.9) r(P, r_0, a)*P*(1-P/K)

@@

$$ C(V) = mV \,, $$

with the grazing mortality, $m$.

```{r, echo=FALSE}
C <- function(V, m = 0.25) (m*V)

@@

$$ C(V) = m \,, $$

thus would be defined as constant, $m$, regardless of the vegetation cover.

```{r, echo=FALSE}
C <- function(V, m = 0.1) m + V*0

@@

@@

$$ C(V) = \frac{aV}{1+ahV} \,,$$

with the handling time, $h$ [time per resource unit], and the search efficiency, $a$ [area foraged in per time].

```{r, echo=FALSE}
C <- function(V, a = 8, h = 10) (a*V)/(1+a*h*V)

@@

@@

@@

```{r, echo=FALSE}
C <- function(V, a = 8, h = 10, q = 1) (a*V^(1+q))/(1+a*h*(V^(1+q)) )

@@

@@

Compared to the more realistic non-linear functions, the constant grazing mortality assumption seems to predict both the attractor of the vegetated state and the tipping point much better than the density dependent mortality used in Kéfi et al 2007 TPB. In fact, the linear density dependent mortality does not predict bistability at all, given a simple logistic growht function.

The critical threshold that would lead to degradation is overestimated by the constant model.

@@

defplot(rho, G(rho), ylab = "growth", col = "green3")
lines(rho, C(rho, a = 10, q = 1))
lines(rho, rep(0.1, length(rho)))
lines(rho, C(rho, a = 3, q = 0))
lines(rho, C(rho, a = .2, h = 0.00001, q = 0))

### open questions

#### Intrinsic mortality

The mortality of plants regardless of grazing cover must be considered to be already internalised into the logistic growth definition.

Alternatively, it can be considered as an independent constant rate, $M(V) = mV$,

$$ \frac{dV}{dt} = G(V) - M(V) - C(V) $$

and thus would affect the alternative stable states of the ecosystem.

defplot(rho, G(rho), ylab = "growth", col = "green3")
m <- rho*0.05
lines(rho, m, lty = 2)
lines(rho, m+C(rho, a = 10, q = 1))
lines(rho, m+rep(0.1, length(rho)))
lines(rho, m+C(rho, a = 3, q = 0))
lines(rho, m+C(rho, a = .2, h = 0.00001, q = 0))

@@

These mechanisms can be implemented in the model by using a pair-approximation approach.

### growth

@@

growth part of equation (28) from @kefi07:

$$  \rho_{0}w_{0,+} = \left( \delta \rho_+ + (1-\delta) \frac{\rho_+ - \rho_{+-} - \rho_{++} }{1 - \rho_+ - \rho_-} \right) (b - c \rho_+)(1- \rho_+ - \rho_-)  $$

The goal of this exercise would be to transform this equation into a function of $\rho_+$, assuming that the landscape is at a vegetated equilibrium.

Therefore, we are looking for the general case of $G(\rho_+)$ where

$$ \frac{d\rho_+}{dt} = \frac{d\rho_-}{dt} = \frac{d\rho_{+,-}}{dt}  = \frac{d\rho_{+,+}}{dt} = \frac{d\rho_{-,-}}{dt}   = 0$$

Equilibrium has to fulfil

$$ \rho_0w_{0,+} = - m \rho_+ \,.$$

thus

$$  - m \rho_+ = \left( \delta \rho_+ + (1-\delta) \frac{\rho_+ - \rho_{+-} - \rho_{++} }{1 - \rho_+ - \rho_-} \right) (b - c \rho_+)(1- \rho_+ - \rho_-)  $$

### spatially-explicit grazing mortality

>>>>

defplot <- function(...) plot(..., 
            type = "l", 
            bty = "l", cex = 0.7, las = 1,
            xlim = c(0,1), ylim = c(0,1), 
            xaxs = "i", yaxs = "i", 
            xaxp = c(0,1,2), yaxp = c(0,1,2), 
            xlab = "vegetation cover")

@@

$$ G(V) = rV(1-\frac{V}{K}) \,,$$

@@

```{r, echo=FALSE}
G <- function(P, r = 2.25, K = 0.9) r*P*(1-P/K)

@@

**This model is going to be the starting point for the model development**

@@

$$ G(V) = r_0 V^{a+1} (1-\frac{V}{K}) \,,$$

@@

G_sig <- function(P, r_0 = 2, a = 2, K = 0.9) r(P, r_0, a)*P*(1-P/K)

defplot(rho, G_sig(rho), ylab = "growth")

@@

$$ C(V) = gVL \,, $$

with the grazing mortality, $g$, per grazer individual multiplied by livestock density, $L$. That is to say, per individual, a constant percentage of cover is eaten. That is, ecologically speaking, not realistic.

```{r, echo=FALSE}
C <- function(V, L = 10, g = 0.01) (g*V*L)

@@

$$ C(V) = gL \,, $$

thus would be defined as constant, $g$, per grazer individual regardless of the vegetation cover.

```{r, echo=FALSE}
C <- function(V, L = 10, g = 0.01) g * L + V*0

@@

Note that grazing mortality can exceed 1, meaning that grazers would exploit more biomass than provided at full cover. Thus stability can be only reached if growth is sufficiently higher and the landscape regenerates itself multiple times per year.

@@

$$ C(V) = \frac{aVL}{1+ahV} \,,$$

with the handling time, $h$ [time per landscape unit], and the search efficiency, $a$ [percent of landscape foraged in per time].

```{r, echo=FALSE}
C <- function(V, L = 10, a = 0.6, h = 100) (a*V*L)/(1+a*h*V)

@@

Since we are calculating in fractions of landscape, and not in cells, the handling time defines the time a grazer individual would need to handle the entire landscape as readily available fodder. The search time is the percentage of the landscape foraged in over the course of one year.

@@

The

@@

```{r, echo=FALSE}
C <- function(V, L = 10, a = 0.6, h = 100, q = 1) (a*V^(1+q)*L)/(1+a*h*(V^(1+q)) )

@@

#### null model

We assume the classic type II functional response plus an intrinsic mortality rate,

$$ C(V) = m*V + \frac{aVL}{1+ahV} \,,$$

C <- function(V, m_0 = 0.05, L = 10, a = 0.6, h = 100, q = 0) (m_0 * V) +(a*V^(1+q)*L)/(1+a*h*(V^(1+q)) )

defplot(rho, C(rho), ylab = "plant mortality")
abline(a = 0, b = 0.05, lty = 2)
```

**This model is going to be the starting point for our model development**

@@

Compared to the more realistic non-linear functions, the constant grazing mortality assumption seems to predict both the attractor of the vegetated state and the tipping point much better than the density dependent mortality used in Kéfi et al 2007 TPB. In fact, the linear density dependent mortality does not predict bistability at all, given a simple logistic growth function.

defplot(rho, G(rho), ylab = "growth", col = "green3")
lines(rho, C(rho, L = 10, a = 0.01, h = 0.00001))
lines(rho, C(rho, L = 25, a = 0.01, h = 0.00001))
lines(rho, C(rho, L = 50, a = 0.01, h = 0.00001))
abline( h = 0.1 )
abline( h = 0.25 )
abline( h = 0.5 )
```

When assuming a type II functional response, the system gains true bistability.

defplot(rho, G(rho), ylab = "growth", col = "green3")
lines(rho, C(rho, L = 10))
lines(rho, C(rho, L = 25))
lines(rho, C(rho, L = 50))

@@

defplot(rho, G(rho), ylab = "growth", col = "green3")
lines(rho, C(rho, L = 10, q = 1))
lines(rho, C(rho, L = 25, q = 1))
lines(rho, C(rho, L = 50, q = 1))

@@

These mechanisms can be implemented in the model by using a pair-approximation approach. 
In numerical simulations $G$ and $C$ can be implemented as the transition rates in a two-state spatial model.

@@

In a model with only global effects, the transition probability of indiviual plant death would be

$$ w_{1,0} = C(V)/V = m + \frac{aL}{1+ahV} $$

and the transition probability of individual plant growth

$$ w_{0,1} = G(V)/V = r (1-\frac{V}{K}) \,.$$

### spatially-explicit growth

We implement the accelerating and decelerating features on a local level.

This is assuming that growth rate is enhanced by a local facilitation term similar to the implementation in Kéfi et al 2007. The growth rate becomes a function of local vegetation cover $\nu$:

$$ r = r_0 + f \nu \,,$$

with an intrinsic growth rate $r_0$ which determines the probability of growth in absence of local vegetation and the additive facilitation $f$ which gradually determines the enhancement of the growth with an increasing local vegetation cover (maximizing at $\nu = 1$ if the cell has 4 neighbors).

#### decelerating growth

This is assuming that growth is diminished locally by competition, $c$. This affects carrying capacity rather than growth rate itself.

$$ K = K_0 (1 - c \nu)$$

### spatially-explicit mortality

On the mortality side, we can implement spatially explicit effects on the parameters of the functional response.

#### accelerating mortality

Mortality becomes enhanced in sparse environments by patches that are attracting grazers. We assume this attractivity, $v$, reduces search times locally.

$$ a = a_0 * (1 - v * \nu) $$

#### decelerating mortality

Plants in grazed habitats develop protective traits, such as thorns or cushion growth. They thereby provide protected habitat to their direct neighborhood, or share the investments in those traits with their neighbors. Overall grazing mortality thus is reduced through an increase in handling time by associational resistance, $p$

$$ h = h_0 * p * \nu $$

$$ $$

#### decelerating mortality
